Reversible Regime Change: Climate-driven phytoplankton community shifts in the Cariaco Basin, Venezuela

### 2025-08-13 agujournaltemplate.tex
backup before final edits before submission

    The calculated genus diversity from the microscopy cell counts generally followed the dynamics of chlorophyll \textit{a} biomass, with maximum values observed during the upwelling seasons in winter. Genus richness was consistently highest in the top \qty{25}{meters} depth interval, also showing the highest variability over the time series. Across the time series, we observed maximum richness at the start of the time series (1996-1997), with a subsequent reduction in diversity across all depths, followed by a recovery after 2011. The reduction in diversity following the start of the time series did not follow the pattern shown by chlorophyll \textit{a}, which remained relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}). Scaling genus richness by the cell count abundances, i.e. the Shannon index, showed a similar but less pronounced dynamic. Diversity remained high throughout 2000 and recovered more quickly but showed marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009, and 2016. The seasonal signal was lost (see Figure \ref{fig:divts} $f$).

@@

\subsection{Clustering analysis yields contrasting ecological regimes}
    % Figure 4
    We undertook clustering analysis on annually aggregated cell counts from microscopy and on individual monthly measurements. Figure \ref{fig:clustering}$a$ shows the results of clustering the annually aggregated communities using a linear complete linkage algorithm based on the binary Jaccard matrix. Two distinct clusters emerge from this analysis, with the first cluster encompassing the years 1996-2004 and 2014-2016, and the second cluster including the intermediate years from 2005 to 2013. 
    Inter-annual species turnover is shown in Figure \ref{fig:clustering}$b$, it reveals an increasing trend in turnover from 1996 to 2004. Turnover subsequently decreased during the period underpinning cluster 2 and increased again in 2014, marking a return to cluster 1. This pattern is also reflected in the hierarchical clustering results, Figure \ref{fig:clustering}$a$, where cluster 2 shows a lower distance in the clustering tree between the years. We additionally extracted the three dominant genera for the second cluster and the early and late parts of the first cluster, respectively. The diatom \textit{Thalassiosira} dominated the early period and was abundant in all other periods. \textit{Pseudo-nitzschia} dominated cluster 2, with the haptophyte \textit{Emiliana} being the second most abundant genus. The later part of cluster 1 is dominated by \textit{Emiliana}.

@@

    % Figure 5 - NEW
    Figure \ref{fig:clustcomp} shows density distributions of environmental variables, chlorophyll \textit{a}, and diversity indices for the two clusters. Upwelling-related variables exhibit distinct patterns between the two clusters. For example, cluster 1 experienced more frequent high wind events, a shallower \qty{21}{\celsius} isotherm, lower temperatures, reduced evaporation, and higher salinity and nutrient concentrations, particularly for nitrate. The AMO index shows a pronounced separation between the two clusters, with cluster 1 exhibiting more negative anomalies, while cluster 2 shows a higher occurrence of positive anomalies. Although there are no significant differences in genus richness and the Shannon index between clusters, there is a slightly increased median in the Pielou index for cluster 2 (Table \ref{tab:ClustCompWilcox}). The density distribution reveals some differences. Cluster 1 shows higher variance for genus richness, whereas cluster 2 shows greater variance and a bimodal distribution for the Shannon and Pielou indices.

>>>>

    The calculated genus diversity from the microscopy cell counts generally followed the dynamics of chlorophyll \textit{a} biomass, with maximum values observed during the upwelling seasons in winter. Genus richness was consistently highest in the top \qty{25}{meters} depth interval, also showing the highest variability over the time series. Across the time series, we observed maximum richness at the start of the time series (1996-1997), with a subsequent reduction in diversity across all depths, followed by a recovery after 2011. The reduction in diversity following the start of the time series did not follow the pattern shown by chlorophyll \textit{a}, which remained relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts} $a$). Scaling genus richness by the cell count abundances, i.e. the Shannon index, showed a similar but less pronounced dynamic. Diversity remained high throughout 2000 and recovered more quickly but showed marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009, and 2016. The seasonal signal was lost (see Figure \ref{fig:divts} $f$).

@@

\subsection{Clustering analysis yields contrasting ecological regimes}
    % Figure 4
    We undertook clustering analysis on annually aggregated cell counts from microscopy and on individual monthly measurements. Figure \ref{fig:clustering}$a$ shows the results of clustering annually aggregated communities using a linear complete linkage algorithm based on the binary Jaccard matrix. Two distinct clusters emerge from this analysis, with the first cluster encompassing the years 1996-2004 and 2014-2016, and the second cluster including the intermediate years from 2005 to 2013. 
    Inter-annual species turnover is shown in Figure \ref{fig:clustering}$b$, revealing an increasing trend in turnover from 1996 to 2004. Turnover subsequently decreased during the period underpinning cluster 2 and increased again in 2014, marking a return to cluster 1. This pattern is also reflected in the hierarchical clustering results, Figure \ref{fig:clustering}$a$, where cluster 2 shows a lower distance in the clustering tree between the years. We additionally extracted the three dominant genera for the second cluster and the early and late parts of the first cluster, respectively. The diatom \textit{Thalassiosira} dominated the early period and was abundant in all other periods. \textit{Pseudo-nitzschia} dominated cluster 2, with the haptophyte \textit{Emiliana} being the second most abundant genus. The later part of cluster 1 is dominated by \textit{Emiliana}.

@@

    % Figure 5 - NEW
    Figure \ref{fig:clustcomp} shows density distributions of environmental variables, chlorophyll \textit{a}, and diversity indices for the two clusters. Upwelling-related variables exhibit distinct patterns between the two clusters. For example, cluster 1 experienced more frequent high wind events, a shallower \qty{21}{\celsius} isotherm, lower temperatures, reduced evaporation, and higher salinity and nutrient concentrations, particularly for nitrate. The AMO index shows a pronounced separation between the two clusters, with cluster 1 exhibiting more negative anomalies, while cluster 2 shows a higher occurrence of positive anomalies. Although there are no significant differences in genus richness and the Shannon index between clusters, there is a slightly increased median in the Pielou index for cluster 2 (Table \ref{tab:ClustCompWilcox}). The density distributions highlight additional contrasts; cluster 1 shows higher variance for genus richness, whereas cluster 2 shows greater variance and a bimodal distribution for the Shannon and Pielou indices.
